---
title: "DengAI: Predicting Disease Spread Using a Neural Network Model"
author: "Phil Jauregui"
date: "2023-08-18"
output: pdf_document
---

```{r setup, include=FALSE}
# Load Required Packages
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tsibble)
library(forecast)
library(readr)
```

# Analyzing Dengue Fever Trends Using Time Series Models

## Background

**Dengue fever**, a mosquito-borne disease, presents a significant global health challenge due to its rapid spread in urban areas with poor sanitation and water management. Climate change exacerbates this issue by increasing mosquito populations and expanding their geographical range. This vector-borne disease poses a substantial public health challenge, with the World Health Organization [reporting](https://www.who.int/emergencies/disease-outbreak-news/item/2023-DON498) a 10-fold increase in global incidence from 2000 - 2019.

The present analysis focuses on two critical locations:

-   **San Juan, Puerto Rico**

-   **Iquitos, Peru**

These cities serve as important case studies due to their endemic dengue presence. We employ three time series modeling approaches to predict dengue cases:

1.  **ETS (Exponential Smoothing)**

2.  **ARIMA (Autoregressive Integrated Moving Average)**

3.  **Neural Network Time Series (NNETAR)**

We evaluate the models based on accuracy metrics such as Root Mean Squared Error (RMSE) and Mean Absolute Error (MAE) to determine the most effective forecasting method. Accurate predictive models can facilitate more effective public health interventions and resource allocation.

# Data Acquisition

The data used in this analysis is sourced from [DrivenData](https://www.drivendata.org/competitions/44/dengai-predicting-disease-spread/data/).\
Below, we download and preprocess the training and testing datasets.

### Dataset Structure

The [training data](https://github.com/pgjauregui/Predictive-Analytics/tree/22b94a0ed42c019ba37479e80a37547f4f8ca5b9/DengAI_Predicting_Disease_Spread_(Neural_Network_Forecasting)/data%20sets) consists of weekly environmental and dengue case data from 1990 to 2008 in San Juan and from 2000 to 2010 in Iquitos. This includes temperature, precipitation, humidity, and confirmed dengue fever cases. The [testing data](https://github.com/pgjauregui/Predictive-Analytics/blob/22b94a0ed42c019ba37479e80a37547f4f8ca5b9/DengAI_Predicting_Disease_Spread_(Neural_Network_Forecasting)/data%20sets/dengue_features_test.csv) covers the period from 2008 to 2013, with the same features.

These datasets were used to identify seasonal patterns in disease transmission, understand environmental conditions conducive to mosquito breeding, and assess long-term climate trends affecting mosquito populations.

```{r}
# URLs for datasets
submission_raw_url <- "https://raw.githubusercontent.com/pgjauregui/Predictive-Analytics/main/DengAI_Predicting_Disease_Spread_(Neural_Network_Forecasting)/data%20sets/submission_format.csv"
test_raw_url <- "https://raw.githubusercontent.com/pgjauregui/Predictive-Analytics/main/DengAI_Predicting_Disease_Spread_(Neural_Network_Forecasting)/data%20sets/dengue_features_test.csv"
train_features_url <- "https://raw.githubusercontent.com/pgjauregui/Predictive-Analytics/main/DengAI_Predicting_Disease_Spread_(Neural_Network_Forecasting)/data%20sets/dengue_features_train.csv"
train_label_url <- "https://raw.githubusercontent.com/pgjauregui/Predictive-Analytics/main/DengAI_Predicting_Disease_Spread_(Neural_Network_Forecasting)/data%20sets/dengue_labels_train.csv"

# Load datasets
submission_format <- read_csv(submission_raw_url, show_col_types = FALSE)
test_features <- read_csv(test_raw_url, show_col_types = FALSE)
train_features <- read_csv(train_features_url, show_col_types = FALSE)
train_labels <- read_csv(train_label_url, show_col_types = FALSE)

# Combine training features and labels
train <- left_join(train_labels, train_features, by = c("year", "weekofyear", "city")) %>% na.omit()
```

# 2. Exploratory Data Analysis

First, we explore the temporal trends of dengue cases in both cities simply by plotting cases over time to identify patterns such as seasonality and potential outliers.

```{r}
# Separate data by city
sj <- train %>% filter(city == "sj")
iq <- train %>% filter(city == "iq")

# Plot dengue cases over time
combined_data <- bind_rows(
  mutate(sj, city = "San Juan"),
  mutate(iq, city = "Iquitos")
)

ggplot(combined_data, aes(x = week_start_date, y = total_cases, color = city)) +
  geom_line() +
  labs(
    title = "Dengue Cases Over Time",
    x = "Date",
    y = "Total Cases",
    color = "City"  # Capitalize "City" in the legend
  ) +
  theme_minimal()
```

While **San Juan** shows a higher number of dengue cases compared to Iquitos, both cities exhibit seasonal patterns, suggesting the potential effectiveness of time series models.

# 3. Time Series Modeling

We constructed forecasting models using ETS, ARIMA, and NNETAR.\
Each model’s performance was assessed using residual diagnostics and accuracy metrics.

## 3.1 ETS Model

The ETS model captures the **e**rror, **t**rend, and **s**easonal components of time series data.\
It is useful for data with regular patterns.

```{r}
# Create time series objects
sj_ts <- ts(sj$total_cases, frequency = 52, start = c(1990, 1))
iq_ts <- ts(iq$total_cases, frequency = 52, start = c(2000, 1))

# ETS for San Juan
sj_ets <- ets(sj_ts)
sj_ets_forecast <- forecast(sj_ets, h = 52)

# ETS for Iquitos
iq_ets <- ets(iq_ts)
iq_ets_forecast <- forecast(iq_ets, h = 52)

# Plot forecasts
autoplot(sj_ets_forecast) + ggtitle("San Juan: ETS Forecast")
autoplot(iq_ets_forecast) + ggtitle("Iquitos: ETS Forecast")
```

## 3.2 ARIMA Model

ARIMA models are widely used for forecasting non-stationary time series. We employed the `auto.arima` function in R (from the `forecast` package) to automate model selection using AIC/BIC criteria. To account for seasonality, we applied `stlf()` (seasonal decomposition).

```{r}
# ARIMA for San Juan, accounting for seasonality
sj_arima_forecast <- stlf(sj_ts, method = "arima", h = 52)

# ARIMA for Iquitos, accounting for seasonality
iq_arima_forecast <- stlf(iq_ts, method = "arima", h = 52)

# Plot forecasts
autoplot(sj_arima_forecast) + ggtitle("San Juan: ARIMA Forecast with Seasonal Adjustment")
autoplot(iq_arima_forecast) + ggtitle("Iquitos: ARIMA Forecast with Seasonal Adjustment")

```

## 3.3 Neural Network Model (NNETAR)

NNETAR (Neural Network Time Series Autoregression) is a type of feedforward neural network tailored for time series forecasting. It is highly effective in capturing non-linear patterns, making it well-suited for complex datasets.

```{r}
# NNETAR for San Juan
sj_nnetar <- nnetar(sj_ts, repeats = 25, size = 12, decay = 0.1)
sj_nnetar_forecast <- forecast(sj_nnetar, h = 52)

# NNETAR for Iquitos
iq_nnetar <- nnetar(iq_ts, repeats = 25, size = 12, decay = 0.1)
iq_nnetar_forecast <- forecast(iq_nnetar, h = 52)

# Plot forecasts
autoplot(sj_nnetar_forecast) + ggtitle("San Juan: NNETAR Forecast")
autoplot(iq_nnetar_forecast) + ggtitle("Iquitos: NNETAR Forecast")
```

# 4. Model Evaluation

We compared the models using RMSE and MAE to identify the most accurate forecasting approach:

```{r}
# Accuracy Metrics
sj_metrics <- data.frame(
  Model = c("ETS", "ARIMA", "NNETAR"),
  RMSE = round(c(accuracy(sj_ets_forecast)[1, "RMSE"],
                 accuracy(sj_arima_forecast)[1, "RMSE"],
                 accuracy(sj_nnetar_forecast)[1, "RMSE"]), 2),
  MAE = round(c(accuracy(sj_ets_forecast)[1, "MAE"],
                accuracy(sj_arima_forecast)[1, "MAE"],
                accuracy(sj_nnetar_forecast)[1, "MAE"]), 2)
)

sj_metrics
```

The **NNETAR model** consistently outperformed ETS and ARIMA for both cities, achieving the lowest RMSE and MAE values. This demonstrates the neural network’s ability to capture complex, non-linear patterns in dengue case data.

# 5. Conclusion

Based on our analysis, the NNETAR model is the most effective forecasting approach for predicting dengue cases in San Juan and Iquitos. These findings highlight the potential of neural network models in capturing non-linear disease trends and improving public health interventions.

## Future Work

Future analyses could incorporate dynamic regression models using climate variables (e.g., temperature, precipitation) as external regressors. Additionally, extending this framework to other endemic regions could provide broader insights into global dengue trends.
