---
title: "Forecasting Temperature (SNAIVE, ARIMA, ETS, Ensemble)"
author: "Phil Jauregui"
date: "`r Sys.Date()`"
output: html_document
---
# 1. Load Packages
```{r}
library(tidyverse)
library(fpp3)
library(tsibble)
library(tsibbledata)
library(lubridate)
library(stats)
library(forecast)
```

# 2. Read in (Global Temp) Data
```{r}
# I am using Global Land Temperatures by State
d <- read.csv("https://query.data.world/s/axgb3nwemqkewpn2xkrv3f36fg2tyi?dws=00000", header=TRUE, stringsAsFactors=TRUE);

# Here, I'm filtering for California
d<-d|>filter(State == "California")
```

# 3. Format as a 5 year time series
```{r}
d <- d %>%
  mutate(Date = yearmonth(as.character(dt))) %>%
  as_tsibble(index = Date) %>%
  filter(year(Date) >= 2006 & year(Date) <= 2010)
# I'm focusing on 5 years of data: 2006 - 2010.
```

```{r}
d <- d[, c(2, 6)]
```

# 4. Plot the overall data

```{r}
# Plotting Overall temperature data over 5 years
d %>%
  autoplot() +
  labs(y = "Average Temperature (Celsius)", x = "Date", title = "Average Temperature in California") +
  geom_smooth(method = "lm", se = FALSE, color = "red")
# Here we see some clear seasonality, but I wanted to zero in to this so I later use the gg_season function

# Plotting Seasonal fluctuation in temperature over the 5 years
d %>% gg_season(AverageTemperature) +
  labs(y = "Average Temperature (Celsius)", x = "Date", title = "Average Temperature in California")
```
As one would expect, we have seasonality wherein temperatures rise in the summer and fall in the winter. 	

Plot an additive decomposition of the data:
```{r}
additivedecomposed <- stl(d, t.window = 12, s.window = "periodic")
plot(additivedecomposed)
```
We can see strong seasonality. 
Somewhat surprisingly, there’s a non-linear trend that doesn't seem to be clearly rising, despite reports of record heatwaves throughout California -- the short time span of this analysis is notable here.

# 5. Identify training and test sets
```{r}
train <-head(d,48)
test <- tail(d, 12)
#  I build a training set based on the first 4 years of data and have the 5th year serve as my test set.
```

# 6. Build SNAIVE, ARIMA, ETS, and Ensemble Models 
```{r}
train.ts <- ts(train$AverageTemperature,frequency = 12, start = c(2006,1))

mySNAIVE=train |>model(Seasonal_Naive_Model=SNAIVE(AverageTemperature))
myarima=auto.arima(train)
myets=ets(train.ts)
myensemble=train |>model(Ensemble_Model=(ETS(AverageTemperature)+ARIMA(AverageTemperature))/2)
# I constructed a seasonal naive, ARIMA, ETS, and ensemble model to forecast the data. 

mypredict1=forecast(mySNAIVE, h=12)
mypredict2=forecast(myarima, 12)
mypredict3=forecast(myets, 12)
mypredict4=forecast(myensemble, h=12)
```


# 7. Evaluate the Models' Accuracy

## SNAIVE
```{r}
evaluation_snaive <- data.frame(
  MAE = mean(abs(mypredict1$.mean - test$AverageTemperature)),
  RMSE = sqrt(mean((mypredict1$.mean - test$AverageTemperature)^2)),
  MAPE = mean(abs((mypredict1$.mean - test$AverageTemperature) / test$AverageTemperature)) * 100)
cat("SNAIVE Model Evaluation:\n")
print(evaluation_snaive)
```

## ARIMA
```{r}
acc_arima <- accuracy(mypredict2, test$AverageTemperature)
print("ARIMA Accuracy:")
print(acc_arima)
checkresiduals(myarima)
```

## ETS
```{r}
acc_ets <- accuracy(mypredict3, test$AverageTemperature)
print("ETS Accuracy:")
print(acc_ets)
checkresiduals(myets)
```

## Ensemble
```{r}
forecast_values <- ts(mypredict4$.mean, frequency = 12, start = c(2010, 1))
actual_values <- ts(test$AverageTemperature, frequency = 12, start = c(2010, 1))
acc_ensemble <- accuracy(forecast_values, actual_values)
print("Ensemble Accuracy:")
print(acc_ensemble)
```

# 8. Plot the Models
ARIMA
```{r}
autoplot(mypredict2)+labs(y="Average Temperature (Celsius)")
```

ETS
```{r}
autoplot(mypredict3)+labs(y="Average Temperature (Celsius)")
```

Overall Models
```{r}
fit=train |>model(ARIMA_Model=ARIMA(AverageTemperature),
                  Seasonal_Naive_Model=SNAIVE(AverageTemperature),
                  ETS_Model=ETS(AverageTemperature),
                  Ensemble_Model=(SNAIVE(AverageTemperature)+ETS(AverageTemperature)+ARIMA(AverageTemperature))/3)

fit |> forecast(test) |> autoplot(test)+labs(y="Average Temperature (Celsius)", title = "Average Temperature in California")
```
Analysis: examining the overall plot, it's difficult to determine which model outperformed the others; but looking at the accuracy metrics, we see that the ensemble model outperformed all others on MAPE (mean absolute percentage error) and MAE (mean absolute error).
Notably, the ETS model outperformed the ensemble model on RMSE (root-mean-square error), but overall it appears the ensemble model has less error. The two models, though, are quite close in the performance.

Deeper analysis and exploration can be found on a [poster] (https://github.com/pgjauregui/Predictive-Analytics/blob/main/Temperature_Forecasting/Temperature%20Forecasting%20Poster%20.pdf) available on my GitHub.