---
title: "Time Series Decomposition"
author: "Phil Jauregui"
date: "`r Sys.Date()`"
output: html_document
---
Below is an exercise in time series decomposition that demonstrates an upward trend in average temperature in California from 1913-2013. 

# 1. Load Packages
```{r}
library(tidyverse)
library(tsibble)
library(tsibbledata)
library(lubridate)
library(fpp3)
```

# 2. Read in (Global Temp) Data
```{r}
# Reading in Global Land Temperatures by State.csv, 
# Available from data.world @ [Global Land Temperatures by State](https://data.world/environmentdata/climate-change-earth-surface/workspace/file?filename=GlobalLandTemperatures%2FGlobalLandTemperaturesByState.csv)
# I've pre-filtered this file for California data
d <- read.csv("https://query.data.world/s/axgb3nwemqkewpn2xkrv3f36fg2tyi?dws=00000", header=TRUE, stringsAsFactors=TRUE);

# Here, I'm filtering for California
d<-d|>filter(State == "California")
```

# 3. Format as a time series
```{r}
filtered_data <- d %>%
  mutate(YearMonth = yearmonth(as.character(dt))) %>%
  as_tsibble(index = YearMonth) %>%
  filter(year(YearMonth) >= 1913 & year(YearMonth) <= 2013)
# I'm focusing on temperature data from 1913 to 2013
```

# 4. Plot the time series
```{r}
filtered_data %>%
  autoplot(AverageTemperature) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(y = "Average Temperature (Celsius)", x = NULL)
```
Here we see a very noisy plot, but there does appear to be a slight upward trend in average temperatures in California (this says nothing if the upward trend is statistically significant).

# 5. Time Series Decomposition

To isolate the components of the time series, I conduct additive time series decomposition:
```{r}
additivedecomposed <- filtered_data %>%
     select(YearMonth, AverageTemperature) %>%
     stl(t.window = 13, s.window = "periodic")
 
plot(additivedecomposed)
```
Here we can more clearly see an upward trend in the 100 year span from 1913-2013.
[I opt to use additive rather than multiplicative decomposition, as additive is recommended if the seasonal component remains similar across all time levels, which appears to be the case in the plot].